---
title: "Class05: Data Viz with GGPlot"
author: "Rachel Felix (PID: A1603764)"
format: gfm
---

Today we are playing with plotting and graphics in R. 

There are lots of ways to make cool figures in R. 
There is "base" R graphics (`plot()`, `his()` `boxplot()` etc.)

There is also add-on packages, like **ggplot**

```{r}
head(cars)
```
Let's plot this with "base" R:

```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let's plot `mpg` vs `disp`
```{r}
plot(mtcars$mpg, mtcars$disp, pch=16)
```


```{r}
hist(mtcars$mpg)
```


## GGPLOT

The main function in the ggplot2 package is **ggplot()**. First I need to install the **ggplot2** package. I can install any package with the function `install.packages()`. 

> **N.B.** I never want to run `instal.packages()` in my quatro source document!!!

```{r}
library(ggplot2)
ggplot(cars) + 
  aes(x=speed, y= dist) +
  geom_point()

```

```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things: 

- The **data** (given with `ggplot(cars)`)
- The **aesthetic** mapping (given with `aes()`)
- The **geom** (given by `geom_point()`)

> For simple canned graphics "base" R is nearly always faster

### Adding more layers

Let's add a line and a title, subtitle and caption as well as custom axis labels 

```{r}
ggplot(cars) + 
  aes(x=speed, y= dist) +
  geom_point() +
  geom_line()
```


```{r}
ggplot(cars) + 
  aes(x=speed, y= dist) +
  geom_point() +
  geom_smooth(method="lm", se= FALSE) +
  labs(title = "Silly Plot", x = "Speed (MPH)", y = "Distance (ft)", caption = "when is tea time anyway???") +
  theme_bw()


```


## Plot some expression data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this dataset? 

There are `r nrow(genes)` in this dataset

> Q2. How many "up" regulated genes are there? 

```{r}
sum(genes$State == "up")
```

There are `r sum(genes$State == "up")` UP genes 

```{r}
table(genes$State)
```

```{r}
round(table(genes$State)/nrow(genes) * 100, 2)
```


## Plot genes

```{r}
ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=Condition2) +
  geom_point() 
  
```

```{r}
p<- ggplot(genes) + 
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point() +
  scale_color_manual(values=c("blue", "grey", "red")) 
```


```{r}
p + labs(title= "look at me")
```


Silly example of adding labels 

```{r}
library(ggrepel)
ggplot(genes) + 
  aes(x=Condition1, 
      y=Condition2, 
      col=State, 
      label=Gene) +
  geom_point() +
  scale_color_manual(values=c("blue", "grey", "red")) + 
  geom_text_repel(max.overlaps = 60) +
  theme_bw()
  
```


## Going Further

Playing with some different layers and the gapmider dataset... 

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```



```{r}
head(gapminder)
```


```{r}
tail(gapminder)
```


A first plot

```{r}
ggplot(gapminder) + 
  aes(y= lifeExp, x=gdpPercap, col=continent) +
  geom_point(alpha = 0.2) +
  facet_wrap(~continent) + 
  theme_bw()
```












